---
title: "16S gene analysis for Mesinkovska skin microbiome project"
author: "Alberto Barron"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, cache = F)
options(scipen=10000, digits = 12)
set.seed(seed = 711)
```

#### Load packages
```{r packages, message=FALSE, warning=FALSE, results='hold'}
## First check that you have these packages installed before trying to load them
library(tidyverse)
library(phyloseq)
library(ggplot2)
library(vegan)
library(dplyr)
library(gplots)
library(RColorBrewer)
library(ggpubr)
library(ggh4x)
library(EcolUtils)
library(biomformat)
library(qiime2R)
```

#### Set global options
- APA theme
- Color palette
- Set working directory
```{r general settings}
knitr::opts_chunk$set(echo = TRUE)

apatheme=theme_bw()+
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        text=element_text(family='Arial'))

taxa_colors <- c(
  "#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD", 
  "#AD6F3B", "#673770","#D14285", "#652926", "#C84248", 
  "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861", "red",
  "blue", "#483D8B", "gray", "darkcyan", "darkgoldenrod4", "green", "#8B0000",
  "darkgreen", "yellow")

setwd("/Users/apple/SyncFolder/skin")

```

### Load data into R

In this step we unzip the files containd in the .qza artifacts, then import OTU table in BIOM format, taxonomy and metadata. Finally we create a 'phyloseq' object that contains the OTU table, taxonomy and metadata.
```{r Load data; pre-porcessing and creating phyloseq object}
## We will unzip our table.qza & taxonomy.qza file.
## By default, qiime2 zips your data, along with provenance info, into a .qza or .qzv file.
## For R to access the data, we need to unzip the files.
## We only need to unzip the files once so after unziping we can comment the next two lines of code.
## Modifiable within quotes, but should work if directory is correct.
#unzip(zipfile = "table1.qza")
## Modifiable within quotes, but should work if directory is correct.
#unzip(zipfile = "taxonomy.qza")

## Now we will import our OTU table by specifying the location of the .biom file, 
## which was obtained from unzipping the table.qza file.
## The OTU table contains the sequence IDs of each OTU and their frequency across your samples.
my_biom<-read_hdf5_biom("b0033878-55ff-4736-8164-4686f4e4ce1c/data/feature-table.biom") ## Tab autocomplete is your friend! Change here within quotes.
write_biom(my_biom, "formatted_biom")
my_biom<-read_biom("formatted_biom")
OTU_table<-as.data.frame(as.matrix(biom_data(my_biom))) %>%           ## Turn our biom file into a dataframe
  rownames_to_column("Feature ID")                                      ## Give the rownames their own column
write_delim(OTU_table, file = "OTU_table_unfiltered.tsv", delim = "\t") ## Save the dataframe to our computer
## Next let's import taxonomy.
## The taxonomy contains the taxa assigned to the OTU sequence ID.
## Location of file modifiable in quotes.

#OTU_taxonomy<-read.table("3059f12d-eb18-4b14-ac4d-2ac29c679002/data/taxonomy.tsv", header = T, sep = "\t")
OTU_taxonomy <- read_qza("taxonomy.qza")

# Parse the taxonomy table to be importable for 'phyloseq'
OTU_taxonomy<-parse_taxonomy(OTU_taxonomy$data)
OTU_taxonomy<-as.matrix(OTU_taxonomy)

## Finally, let's import the metadata, this should be writen in the working directory as a .txt file with tabs a separators.
metadata<-read.table("metadata.txt", header = TRUE, sep = '\t', row.names = 1)

#get sample names in metadata
metadata_sample_names<-row.names(metadata)
#get sample names in OTU table
OTUtab_sample_names<-colnames(OTU_table)
OTUtab_sample_names<-OTUtab_sample_names[-1]
#get shared sample names in metadata and OTU table
shared_samples<-base::intersect(metadata_sample_names, OTUtab_sample_names)
#get sample names in the OTU table NOT present in the metadata
not_shared<-setdiff(OTUtab_sample_names, metadata_sample_names)
#get sample names in the the metadata NOT present in OTU table
missing_in_md<-setdiff(metadata_sample_names, OTUtab_sample_names)

## Now let's remove the duplicated samples from the OTU table. These are all the columns whose name ends with "-2"
## First let's create another table with those samples
# get the sample IDs of the duplicates
dup_ids<-sub(".{2}$", "", not_shared)
#Create a vector with the columns to keep
names.keep <- names(OTU_table)[(names(OTU_table) %in% dup_ids)]
#subset OTU table
OTU_tab_duplicates<-OTU_table[, names.keep]
#add "Feature ID" column and move it to the first column
OTU_tab_duplicates$`Feature ID`<-OTU_table$`Feature ID`
OTU_tab_duplicates<-OTU_tab_duplicates %>% relocate(`Feature ID`)
#get sample sums
sampsums_dup<-colSums(OTU_tab_duplicates[,-1])


# Let's compare the duplicate samples to know which ones to include

#Create a vector with the columns to keep
names.keep.dup <- names(OTU_table)[(names(OTU_table) %in% not_shared)]
#subset OTU table
OTU_tab_duplicates_1<-OTU_table[, names.keep.dup]
#add "Feature ID" column and move it to the first column
OTU_tab_duplicates_1$`Feature ID`<-OTU_table$`Feature ID`
OTU_tab_duplicates_1<-OTU_tab_duplicates_1 %>% relocate(`Feature ID`)
#get sample sums
sampsums_dup1<-colSums(OTU_tab_duplicates_1[,-1])

sampsums_dup_summary<-as.data.frame(cbind(sampsums_dup, sampsums_dup1))
colSums(sampsums_dup_summary)
```

Sequence duplicates have more samples than the first ones, let's exclude the duplicate samples from the OTU table

```{r include duplicates and remove original ones}
#Create a vector with the columns to exclude
names.exclude <- names(OTU_table)[!(names(OTU_table) %in% dup_ids)]
#subset OTU table
OTU_tab_filtered<-OTU_table[, names.exclude]

# Rename samples to the original id to match the sample id in the metadata
OTU_tab_filtered<-OTU_tab_filtered %>% rename_at(vars(not_shared), ~ dup_ids)

## Now let's subset the metadata to remove the two samples not present in the OTU table
metadata<-metadata[!rownames(metadata) %in% missing_in_md, ]
## Confirm that the only difference between row names in the metadata and column names in the OTU table is the "Feature ID" column
base::setdiff(colnames(OTU_tab_filtered), row.names(metadata))

## Now let's create the phyloseq object with the OTU table, taxonomy and metadata.
# Set column "Feature ID" as row names
row.names(OTU_tab_filtered)<-OTU_tab_filtered$`Feature ID`
# Remove first columns from OTU table as it has the Feature ID which now are the row names
OTU_tab_filtered<-OTU_tab_filtered[,-1]
#Create 'phyloseq' importable objects
OTUtab<-otu_table(OTU_tab_filtered, taxa_are_rows = T)
tax<-tax_table(OTU_taxonomy)
sam_data<-sample_data(metadata)
#create 'phyloseq' object and check that metadata and taxonomy are imported correctly
dat<-phyloseq(OTUtab, tax, sam_data)
dat
```
### Creating new variables in the metadata with the detailed metadata from each subject

Before continuing with the processing of the OTU table and diversity analysis we need to match some information about the subjects contained in the detailed metadata to their corresponding subject in the OTU table metadata.

```{r processing matadata}
## First let's remove uninformative collumns in the metadata (columns with technical information about the library preparation)
dat@sam_data<-dat@sam_data[,-c(5:16)]
# Get summary of the metadata
summary(dat@sam_data)
# Assign categorical variables
dat@sam_data$subject<-factor(dat@sam_data$subject)
dat@sam_data$Timepoint<-factor(dat@sam_data$Timepoint)
dat@sam_data$Treatment<-factor(dat@sam_data$Treatment)

## Now, let's load the detailed metadata
det_md<-read.table("detailed_metadata.txt", header = TRUE, sep = '\t', colClasses = "factor")

dat@sam_data$smplename<-row.names(as.data.frame(as.matrix(dat@sam_data)))

mdf<-merge(as.data.frame(as.matrix(dat@sam_data)), det_md, by.x="subject", by.y="SubjectID", all.x=TRUE) %>%
#mdf <- mdf %>% 
  mutate(Asthma = recode(Asthma, no = "No")) %>%
  mutate(Food_allergies = recode(Food_allergies, no = "No")) %>%
  mutate(Family_history_Asthma = recode(Family_history_Asthma, no = "No")) %>%
  mutate(Family_history_allergic_rhinitis = recode(Family_history_allergic_rhinitis, yes = "Yes")) %>%
  mutate(Family_history_food_allergies = recode(Family_history_food_allergies, no = "No")) %>%
  mutate(Family_history_AD = recode(Family_history_AD, yes = "Yes")) %>%
  column_to_rownames("smplename")

summary(mdf)
dat@sam_data<-sample_data(mdf)

## Finally check that all new variables had been assigned correctly and coded as factorial variable
summary(dat@sam_data)
```

### Rarefaction curves

Now that we have formatted the metadata and added informative variables, let's see create some rarefaction curves to visualize ASV diversity as a function of the number of sequences in each sample. For this we will use the 'rarecurve' function od the package 'vegan' to calculate the rarefaction curves. A dashed red line is drawn at the rarefaction depth that we have chosen, **3251 sequences**.

```{r Rarefaction curves, fig.width=12}
tab<-otu_table(dat)
class(tab)<-"matrix"
tab<-t(tab)

rar_curve_df <- rarecurve(tab, step = 1000, label = F, sample = 3514, xlab = "Read Depth", col = "orange", tidy = TRUE)

rar_curves<- rar_curve_df %>%
  filter(!Site %in% c("lysis-buffer", "lysis-buffer-2", "Mock", "Mock2",  "negative", "negative2")) %>%
  ggplot(aes(x=Sample, y=Species, group = Site)) + 
  geom_line() +
  geom_vline(xintercept = 3251, color = "red", linetype = "dashed") +
  #ggbreak::scale_x_cut(c(60000), which = c(2), scales = c(0.5, 2)) +
  labs(x="Number of sequences", y="Number of ASVs") +
  guides(fill="none") +
  apatheme +
  theme(legend.position = "bottom", 
        legend.title=element_blank(),
        text=element_text(size=12)
        #,
        #axis.text.x = element_text(angle=90, hjust=1, vjust=.5)
        )
print(rar_curves)

ggsave("Rarefaction_curve.svg",
       width = 25,
       height = 20,
       units = "cm",
       dpi = 600)


```

### Pre-processing for diversity analysis

#### Removing undesired sequences

Before proceeding with the diversity analysis we have to remove eukaryotic sequences (Chloroplasts and Mitochondria), sequences not classified and finally rarefy the OTU table.

```{r  sequence count and removing eukaryotic sequences}

#identify and count plant-derived sequences
data_chll<-subset_taxa(dat, Genus == "Chloroplast")
ntaxa(data_chll)
chl_sums<-sum(sample_sums(data_chll))
chl_sums

data_mit<-subset_taxa(dat, Genus == "Mitochondria")
ntaxa(data_mit)
mit_sums<-sum(sample_sums(data_mit))
mit_sums

data_euk<-subset_taxa(dat, Kingdom == "d__Eukaryota")
ntaxa(data_euk)
euk_sums<-sum(sample_sums(data_euk))
euk_sums

euk_derived_seqs<-chl_sums + mit_sums + euk_sums
euk_derived_seqs

#Check total sequences count before removing Eukaryota derived sequences
total_seqs<-sum(sample_sums(dat))
total_seqs

#calculate % of plant-derived seqs
(euk_derived_seqs/total_seqs)*100

#remove Chloroplast & Mitochondria identified OTUs and unassigned OTUs at the Kingdom level
dat<-subset_taxa(dat, (Genus!="Chloroplast" | is.na(Genus)))
ntaxa(dat)

dat<-subset_taxa(dat, (Genus!="Mitochondria" | is.na(Genus)))
ntaxa(dat)

dat<-subset_taxa(dat, (Kingdom!="d__Eukaryota" | is.na(Kingdom)))
ntaxa(dat)

dat<-subset_taxa(dat, (Kingdom!="Unassigned" | is.na(Kingdom)))
ntaxa(dat)

# Total sequences after removing undesired sequences 
sum(sample_sums(dat))

## Check for the lowest sequence count
min(sample_sums(dat))          #check lowest sequence count to rarefy
#[1] 2655                       #minimum sequence count too low

## Let's check how the sequencing and extraction controls look.
## Let's create a table that contains sequence count per sample and the gel result information from the metadata
sampsums<-as.data.frame(sort(sample_sums(dat)))
sampsums$sample_id<-row.names(sampsums)
sampsums$gel_result<-metadata$gel_result_16S[match(sampsums$sample_id, row.names(metadata))]
names(sampsums)[1]<-"count"

as.data.frame((head(sampsums[,-2])))
```

Lowest sequence count is a negative control, we will rarefy to the next lowest sequences count, **3251** .

```{r, fig.width=12}
## let's check how many samples have a "not present" gel result on the metadata
table(sampsums$gel_result)
# 34 samples have a "not present" gel result

# Now let's check how many samples have a "not present" gel result and have a sequence count below the sequence count of the highest sequence count in the controls
# lysis-buffer: sequence count = 16120
length(which(sampsums$count < 16113 & sampsums$gel_result == "not present"))
# 15 samples fulfill both criteria

# Now let's get the name of these samples, since we will exclude these samples from further analysis
samps_to_rm<-row.names(sampsums[which(sampsums$count < 16113 & sampsums$gel_result == "not present"),])
samps_to_rm

## Now let's plot the sequence count
ggplot(data = sampsums, aes(x = reorder(sample_id, count), y = count)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample Id", y = "Sequence count") +
  geom_hline(yintercept = 3251, color = "red", linetype = "dashed") +
  ggbreak::scale_y_cut(c(50000), which = c(2), scales = c(3, 0.5)) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave("seq_count_by_sample.svg",
       width = 50,
       height = 25,
       units = "cm",
       dpi = 600)
```

From the sequence count and the gel result information we can see that the highest sequence count in the control samples is the lysis-buffer  (16120 sequences). We will rarefy to the lowest sequence count in a sample (3251). We will also remove all the samples that have a sequence count below lysis-buffer  (16113 sequences) **AND** have a "not present" gel result (15 samples).

### Rarefaction of OTU table

```{r Rarefaction, message=FALSE, warning=FALSE, results='hide'}

rar_table<-data.frame(t(rrarefy.perm(data.frame(t(dat@otu_table)),
                                     sample = 3251, n = 100,
                                     round.out = T)))
rar_table<-rar_table[rowSums(rar_table)!=0,]
colnames(rar_table)<-gsub("\\.", "-", colnames(rar_table))
colnames(rar_table)<-sub("X", "", colnames(rar_table))
rownames(rar_table)<-sub("X", "", rownames(rar_table))

rar_data<-phyloseq(sample_data(dat@sam_data),
                   otu_table(as.matrix(rar_table), taxa_are_rows = T),
                   tax_table(dat@tax_table))

rar_data

#check for sample sums in rarefied table
sort(sample_sums(rar_data))

## Now, let's remove the samples we decided to exclude

rar_data<-prune_samples(!(sample_names(rar_data) %in% samps_to_rm), rar_data)
rar_data
```

Now let's count how many ASVs and Samples we have after rarefaction.

```{r counting ASVs and samples before and after rarefaction}
## Number of ASVs before rarefaction
length(taxa_names(dat))

## Number of ASVs after rarefaction
length(taxa_names(rar_data))

## Number of samples before rarefaction
length(row.names(dat@sam_data))

## Number of samples after rarefaction
length(row.names(rar_data@sam_data))

```

#### Counting samples in each group

```{r Counting samples}

## Counting samples by subject
rar_data@sam_data %>%
  group_by(subject) %>%
  tally() %>%
  print(n = 23)

## Counting samples by treatment
rar_data@sam_data %>%
  group_by(Treatment) %>%
  tally()

## Counting samples by timepoint
rar_data@sam_data %>%
  group_by(Timepoint) %>%
  tally()

## Counting samples by subject and treatment
rar_data@sam_data %>%
  group_by(subject, Treatment) %>%
  tally() %>%
  print(n = 43)

## Counting samples by Treatment and Timepoint
rar_data@sam_data %>%
  group_by(Timepoint, Treatment) %>%
  tally()

## Counting samples by Gender
rar_data@sam_data %>%
  group_by(Gender) %>%
  tally()

## Counting samples by Gender and Treatment
rar_data@sam_data %>%
  group_by(Gender, Treatment) %>%
  tally()

## Counting samples by Race
rar_data@sam_data %>%
  group_by(Race) %>%
  tally()

## Counting samples by Gender and Race
rar_data@sam_data %>%
  group_by(Gender, Race) %>%
  tally()

## Counting samples by Asthma diagnostic 
rar_data@sam_data %>%
  group_by(Asthma) %>%
  tally()

## Counting samples by Rhinitis diagnostic 
rar_data@sam_data %>%
  group_by(Allergic_Rhinitis) %>%
  tally()

## Counting samples by Rhinitis diagnostic and Treatment
rar_data@sam_data %>%
  group_by(Allergic_Rhinitis, Treatment) %>%
  tally()

## Counting samples by Food Allergies
rar_data@sam_data %>%
  group_by(Food_allergies) %>%
  tally()

## Counting samples by Family Asthma history
rar_data@sam_data %>%
  group_by(Family_history_Asthma) %>%
  tally()

## Counting samples by Family Rhinitis history
rar_data@sam_data %>%
  group_by(Family_history_allergic_rhinitis) %>%
  tally()

## Counting samples by Family Rhinitis history and Treatment
rar_data@sam_data %>%
  group_by(Family_history_allergic_rhinitis, Treatment) %>%
  tally()

## Counting samples by Family food allergies history
rar_data@sam_data %>%
  group_by(Family_history_food_allergies) %>%
  tally()

## Counting samples by Family history of AD
rar_data@sam_data %>%
  group_by(Family_history_AD) %>%
  tally()
```

## Diversity analysis

### Alpha Diversity

Now let's proceed with the alpha diversity analysis. For this we will calculate the Shannon diversity index and a few other indexes used to measure alpha diversity.

#### Shannon-Weinner diversity index

```{r Alpha diversity analyisis - Shannon index}
# Calculate alpha diversity analysis and reformat sample names
divdf<-estimate_richness(rar_data, measures = c("Observed", "Chao1", "Shannon", "Simpson", "InvSimpson"))
rownames(divdf)<-gsub("\\.", "-", row.names(divdf))
rownames(divdf)<-sub("X", "", row.names(divdf))
# Add metadata information to the newly created data frame with the calculated values
divdf<-cbind(divdf, rar_data@sam_data)
```

##### Shannon diversity values per treatment

```{r Shannon diversity values per treatment}
div_treatment<-ggplot(data = divdf, aes(x = Treatment, y = Shannon, fill = Treatment)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  apatheme

print(div_treatment)

ggsave("shannon_by_treatment.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)

## t.test to compare Shannon diversity values between treated and untreated

#ƒ First let's check for normal distribution of Shannon diversity values
shapiro.test(divdf$Shannon)
# For each group
shapiro.test(divdf[which(divdf$Treatment == "T"), 4])
shapiro.test(divdf[which(divdf$Treatment == "U"), 4])

## Shannon diversity is has a normal distribution let's proceed with the t test
divdf_nc <- divdf %>% filter(Timepoint != "na")

t.test(Shannon ~ Treatment, data = divdf_nc)

```
No significant differences between Treated and Untreated samples in terms of Shannon index values.

##### Shannon diversity values per treatment and subjects

```{r Shannon diversity values per treatment and subjects}

div_treat_sub<-ggplot(data = divdf, aes(x = subject, y = Shannon, fill = Treatment)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  apatheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(div_treat_sub)

ggsave("shannon_by_treatmentANDsubject.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)

## Now let's test for an interactive effect between treatment and subject with a 2-way ANOVA

car::Anova(aov(Shannon ~ Treatment*subject, data = divdf_nc), type = "III")
```
When testing the interaction between Treatment and subject with a 2-way ANOVA only the 'subject' factor is significant and the interaction is not significant.


##### Shannon diversity values per treatment, time point and subject

```{r Shannon diversity values per treatment time point and subject}
div_treat_sub_time<-ggplot(data = divdf,
                           aes(x = Timepoint, y = Shannon,
                               color = Treatment,
                               group = Treatment,
                               shape = Treatment)) +
                      geom_point() +
                      geom_line() +
                      facet_wrap(~subject) +
                      apatheme

print(div_treat_sub_time)

ggsave("shannon_by_treatmentANDsubjectANDtimepoint.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)
```

##### Shannon diversity means +/- sd per treatment and time point

```{r Shannon diversity means +/- sd per treatment and time point}
## Now, let's plot Shannon diversity means +/- sd per treatment and time point
# Before that we have to calculate the mean and sd for each time point and treatment
divsum<-divdf %>%
  dplyr::group_by(Timepoint, Treatment) %>%
  dplyr::summarize(avg = mean(Shannon),
                   n = dplyr::n(),
                   sd = sd(Shannon))
## Now we can plot the means and sd per time point and treatment
divsum_treat_time<-ggplot(data = divsum,
                           aes(x = Timepoint, y = avg,
                               color = Treatment,
                               group = Treatment,
                               shape = Treatment)) +
                      geom_point() +
                      geom_line() +
                      geom_errorbar(aes(ymin = avg-sd, ymax = avg+sd), width = .2) +
                      ylab("Shannon") +
                      apatheme

print(divsum_treat_time)

ggsave("shannon_by_treatmentANDtimepoint_means.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)

# Finally let's test the interaction between Timepoint ans treatment
car::Anova(aov(Shannon ~ Treatment*Timepoint, data = divdf_nc), type = "III")
```
No significant effect of Treatment or Time point on the Shannon index values.

#### Observed richness

##### Observed OTUs values per treatment

```{r Alpha diversity analyisis - Observed richness}
divO_treatment<-ggplot(data = divdf, aes(x = Treatment, y = Observed, fill = Treatment)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  apatheme

print(divO_treatment)

ggsave("richness_by_treatment.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)

## t.test to compare Observed richness values between treated and untreated

# First let's check for normal distribution of Observed richness values
shapiro.test(divdf$Observed)
# For each group
shapiro.test(divdf[which(divdf$Treatment == "T"), 1])
shapiro.test(divdf[which(divdf$Treatment == "U"), 1])

# Observed richness doesn't have a normal distribution let's log transform it and see how it looks
shapiro.test(log(divdf$Observed))
# For each group
shapiro.test(log(divdf[which(divdf$Treatment == "T"), 1]))
shapiro.test(log(divdf[which(divdf$Treatment == "U"), 1])) 

##  Log transformed Observed richness has a normal distribution let's proceed with the t.test


t.test(log(Observed) ~ Treatment, data = divdf_nc)

```

No significant differences between Treated and Untreated samples in terms of log transformed Observed richness values.

##### Observed OTUs per treatment and subjects

```{r Observed OTUs per treatment and subjects}
divO_treat_sub<-ggplot(data = divdf, aes(x = subject, y = Observed, fill = Treatment)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  apatheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(divO_treat_sub)

ggsave("richness_by_treatmentANDsubject.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)

car::Anova(aov(log(Observed) ~ Treatment*subject, data = divdf_nc), type = "III")
```

When testing the interaction between Treatment and subject with a 2-way ANOVA only the 'subject' factor is significant and the interaction is not significant.

##### Observed OTUs per treatment, time point and subjects

```{r Observed OTUs per treatment timepoint and subject}
divO_treat_sub_time<-ggplot(data = divdf,
                           aes(x = Timepoint, y = Observed,
                               color = Treatment,
                               group = Treatment,
                               shape = Treatment)) +
                      geom_point() +
                      geom_line() +
                      facet_wrap(~subject) +
                      apatheme

print(divO_treat_sub_time)

ggsave("richness_by_treatmentANDsubjectANDtimepoint.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)
```

##### Observed OTUs means +/- sd per treatment and time point

```{r Observed OTUs means +/- sd per treatment and time point}
## Now, let's plot Shannon diversity means +/- sd per treatment and time point
# Before that we have to calculate the mean and sd for each time point and treatment
divsumO<-divdf %>%
  dplyr::group_by(Timepoint, Treatment) %>%
  dplyr::summarize(avg = mean(Observed),
                   n = dplyr::n(),
                   sd = sd(Observed))
## Now we can plot the means and sd per tiempoint and treatment
divsumO_treat_time<-ggplot(data = divsumO,
                           aes(x = Timepoint, y = avg,
                               color = Treatment,
                               group = Treatment,
                               shape = Treatment)) +
                      geom_point() +
                      geom_line() +
                      geom_errorbar(aes(ymin = avg-sd,
                                        ymax = avg+sd),
                                    width = .2) +
                      ylab("Observed") +
                      apatheme

print(divsumO_treat_time)

ggsave("richness_by_treatmentANDtimepoint_means.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)

# Finally let's test the interaction between Timepoint ans treatment
car::Anova(aov(log(Observed) ~ Treatment*Timepoint, data = divdf_nc), type = "III")
```
No significant effect of Treatment or Time point on the log transformed Observed richness values.

### Taxa Barplots

#### Genus level

Taxa bar plot at the genus level. OTUs with < 0.75% abundance excluded.

```{r taxa bar plots, fig.width=12}
genus_data<-rar_data %>%
  tax_glom(taxrank = "Genus")#agglomerate at genus level
phy_g<-transform_sample_counts(genus_data, function(x) {x/sum(x)} )  #transform to relative abundance
phy_gf<-filter_taxa(phy_g, function(x) mean(x) < 0.0075, TRUE) #filter out Phyla with < 1% abundance
rmtaxa<-taxa_names(phy_gf)
alltaxa<-taxa_names(phy_g)
mytaxa<-alltaxa[!alltaxa %in% rmtaxa]

phy_gf_0.01<-prune_taxa(mytaxa, phy_g) #prune phyla below 0.75% in each sample

phy_gf_0.01<-transform_sample_counts(phy_gf_0.01, function(x) {x/sum(x)} ) %>% #transform to relative abundance
  psmelt() %>% #melt to long format
  arrange(Genus) #sort data frame alphabetically by genus
phy_gf_0.01$Timepoint<-factor(phy_gf_0.01$Timepoint, levels = c("V2", "V3", "V4", "V5", "V6", "na"))

genus_names<-sort(unique(phy_gf_0.01$Genus))
genus_names[2]<-"Rhizobium"

sample_labels<-phy_gf@sam_data[with(phy_gf@sam_data, order(subject, Treatment, Timepoint)),]

#make plot
#phy_gf_0.01 %>%
 # mutate(Sample = factor(Sample, levels = row.names(sample_labels))) %>%
  ggplot(phy_gf_0.01[!(phy_gf_0.01$Timepoint == "na"),],
                      aes(x = Sample, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = taxa_colors, labels = genus_names) +
  #theme_bw() +
  facet_nested(~ subject + Treatment, scale = "free", space = "free_x") +
  #facet_grid(~Treatment, scale = "free", space = "free_x") +
  xlab("Timepoints") +
  scale_x_discrete(labels = sample_labels$Timepoint) +
  scale_y_continuous(labels = scales::percent) + ylab("Abundance (%)") +
  theme_bw() +
  theme(panel.spacing = unit(0.125,"line"),
        panel.border = element_rect(color = "gray", fill = NA, linewidth=0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle=90, hjust=1)
        #axis.text.x = element_blank()
        )

#print(barplot_genus)

ggsave("taxa_barplot_genus.svg",
       width = 50,
       height = 25,
       units = "cm",
       dpi = 600)

```

### Beta diversity analysis

Finally, let's compare the composition of the samples across the different experimental factors

```{r remove control samples}
## Before comparing the composition of the communities we have to subset the control samples so they don't obscure the analysis

rar_data<-subset_samples(rar_data, Timepoint!="na")
rar_data
```

###### Exporting Bray-Curtis distance matrix to perfom PERMANOVAS in PRIMER+ (this is the way)

```{r Export data for PRIMER+}
## Export distance matrix to import into PRIMER with associated metadata

## This is meant to be run ONLY ONCE to export the matrix and metadata to .csv files, that's why it's commented 
#OTU_bray<-as.data.frame(as.matrix(vegdist(as.data.frame(t(rar_data@otu_table)), method = "bray")))
#write.csv(OTU_bray, "bray_matrix.csv", quote = FALSE)
#md4pr<-data.frame(rar_data@sam_data)
#md4pr<-md4pr %>% dplyr::mutate(GenderBin = ifelse(Gender == "M", 0, 1)) %>%
#  dplyr::mutate(Hispanic.LatinoBin = ifelse(Hispanic.Latino == "Yes", 1, 0)) %>%
#  dplyr::mutate(AsthmaBin = ifelse(Asthma == "Yes", 1, 0)) %>%
#  dplyr::mutate(RhinitisBin = ifelse(Rhinitis == "Yes", 1, 0)) %>%
#  dplyr::mutate(FoodAllergiesBin = ifelse(Food_allergies == "Yes", 1, 0)) %>%
#  dplyr::mutate(FammilyAsthmaBin = ifelse(Family_asthma == "Yes", 1, 0)) %>%
#  dplyr::mutate(FamilyRhinitisBin = ifelse(Family_rhinitis == "Yes", 1, 0)) %>%
#  dplyr::mutate(Fammily_FoodAllergiesBin = ifelse(Family_Food_allergies == "Yes", 1, 0)) %>%
#  dplyr::mutate(Family_ADBin = ifelse(Family_AD == "Yes", 1, 0)) %>%
#  dplyr::mutate(TimepointNum = ifelse(Timepoint == "V2", 2,
#                                                     ifelse(Timepoint == "V3", 3,
#                                                            ifelse(Timepoint == "V4", 4,
#                                                                   ifelse(Timepoint == "V5", 5, 6)))))
#write.csv(md4pr, "bray_metadata.csv", quote = FALSE)

```

##### NMDS plot by Treatment

```{r calculate NMDS, results='hide'}
nmds<-ordinate(rar_data,
               method = "NMDS",
               distance = "bray",
               weighted = TRUE)
nmds_df<-plot_ordination(rar_data, nmds, justDF = TRUE)
```
```{r NMDS by Treatment}
nmds_plot<-ggplot() +
  geom_point(data = nmds_df,aes(x=NMDS1, y=NMDS2,
                                color = Treatment,
                                ),
             size = 4, alpha = 0.75, show.legend = TRUE) +
  scale_color_manual(values = taxa_colors) +
  annotate("text", x = -0.4 , y = -0.2, label = paste("Stress =",round(nmds$stress, digits = 3)), size = 3.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

print(nmds_plot)

ggsave("nmds_by_treatment.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)
```

##### NMDS plot by Subject and Treatment

```{r NMDS plot by Subject, fig.width=8, fig.height=8}
nmds_plot_sub<-ggplot() +
  geom_point(data = nmds_df,aes(x=NMDS1, y=NMDS2,
                                color = subject,
                                shape = Treatment
                                ),
             size = 4, alpha = 0.75, show.legend = TRUE) +
  scale_color_manual(values = taxa_colors) +
  theme_bw() +
  annotate("text", x = -0.4 , y = -0.2, label = paste("Stress =",round(nmds$stress, digits = 3)), size = 3.5) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

print(nmds_plot_sub)

ggsave("nmds_by_subject.svg",
       width = 20,
       height = 20,
       units = "cm",
       dpi = 600)
```

#### NMDS plot by treatment and faceted by Subject
```{r NMDS plot by treatment and faceted by Subject}
nmds_plot_sub_fac<-ggplot() +
  geom_point(data = nmds_df,aes(x=NMDS1, y=NMDS2,
                                color = Treatment,
                                #shape = Treatment
                                ),
             size = 2.5, alpha = 0.75, show.legend = TRUE) +
  scale_color_manual(values = taxa_colors) +
  geom_text(data = nmds_df, #add sample ID labels
            aes(x=NMDS1, y=NMDS2-0.005, label=Timepoint), size=2.5, show.legend = F) +
  #annotate("text", x = -0.17 , y = -0.12, label = "Stress = 0.227", size = 2.5) +
  facet_wrap(~subject) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

print(nmds_plot_sub_fac)

ggsave("nmds_by_treatment_facetby_subject.svg",
       width = 35,
       height = 30,
       units = "cm",
       dpi = 600)
```


